Lossy compression of DNA sequencing quality data

### 2015-04-21 manuscript/q-compression.tex
Added section on extending polynomial regression.  Text clean up.

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

@@

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

>>>>

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. All compression methods have been losslessly compressed with \textsc{bz}ip2. Asterisks denote the corresponding lossless compression of the original data using
  \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

It is important to highlight the result that the 2-bin method aligns a greater portion of reads than the various regression models.
If a poor quality base is flanked by high quality sequence, then the low degree polynomial regression models tend to smooth out the base quality, erroneously marking the low quality base as higher quality.
During alignment, Bowtie2 penalizes mismatches at high quality bases more so than mismatches at lower quality bases.
Thus, the polynomial regression models incur a high penalty for these bases, resulting in less alignments, despite having a better \textsc{mse} than 2-bin in most cases.

@@

\subsection{Extension of polynomial regression}

The downside of modeling quality sequences using polynomial regression is that the model often requires a high number of degrees to achieve the same \textsc{mse} as the profile and \textsc{QualComp} methods.
However, storing a high number of coefficients requires more storage than losslessly compressing the original data.
In order to increase the compressibility of modeling, we can attempt to store the \emph{profiles} of certain polynomial functions.
In other words, we can use a spline (a function that is piecewise-defined by polynomial functions) to represent a given quality sequence.
Similar to our profile-encoding method, the user can specify how many polynomial functions they wish to store.
Then a quality sequence can be divided evenly into a given number of segments, and each segment can be annotated with a polynomial function profile that closely matches its quality sequence.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using two
high-coverage bacterial, human, and mouse data sets.
A more direct comparison against additional lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
performed. Furthermore, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.
